---
title: "Getting started with zuhtml"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with zuhtml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

zuhtml turns real-world HTML into ordinary R values: character vectors,
lists and data frames. It parses the way a browser does, so malformed
markup is repaired rather than rejected, and it never fetches anything:
you give it a string, raw bytes or a file.

```{r}
library(zuhtml)
```

## A page to work with

A small catalogue page, as it might have been saved from a site. Some
markup is sloppy on purpose: unclosed `<li>`s, unquoted attributes, a
stray end tag, and a product card without a price.

```{r}
page <- '
<!DOCTYPE html>
<title>Tea shop</title>
<nav><ul><li><a href="/">Home</a><li><a href="sale/">Sale</a></ul></nav>
<p>Free shipping over 30 EUR</span>
<div class=product>
  <h2 class=name>Sencha</h2><span class=price>3.50</span>
  <a href="sencha.html">details</a>
</div>
<div class=product>
  <h2 class=name>Genmaicha</h2>
  <a href="genmaicha.html">details</a>
</div>
<table>
  <thead><tr><th>Size<th>Grams</thead>
  <tr><td>Small<td>0100
  <tr><td>Large<td>0250
</table>'

doc <- html_parse(page, base_url = "https://example.org/shop/")
doc
```

`html_read()` does the same for a file. `base_url` is where the page came
from; relative links are resolved against it.

## Selecting elements

`html_elements()` finds every element that matches a CSS selector.
`html_element()` finds the first match below *each* input node, and keeps
a missing node where there is none. That is what keeps extracted columns
aligned when some records lack a field:

```{r}
cards <- html_elements(doc, ".product")
cards

products <- data.frame(
  name = html_text_clean(html_element(cards, ".name")),
  price = html_text_clean(html_element(cards, ".price")),
  url = html_url(html_element(cards, "a"))
)
products
```

Genmaicha has no price, so it gets `NA` rather than shifting the column.

## Values

`html_text_clean()` gives text as a reader wants it; `html_text()` gives
it exactly as parsed. `html_attr()` reads attributes, and
`html_serialize()` writes nodes back as HTML.

```{r}
html_text_clean(html_element(doc, "title"))
html_attr(html_elements(doc, "nav a"), "href")
html_serialize(html_element(doc, "h2"))
```

## Structures

Links, lists and tables have their own extractors:

```{r}
html_links(doc, absolute = TRUE)
lapply(html_elements(doc, "nav ul"), html_list)
html_tables(doc)
```

Table columns are character: `"0100"` keeps its leading zero. Convert
types yourself when you know them, for example with `type.convert()`.

## What the parser repaired

Real pages nearly always have markup errors, which the parser repairs.
`html_problems()` lists them:

```{r}
html_problems(doc)
```

## Where next

* `vignette("selectors")`: the supported CSS subset.
* `vignette("tables-and-lists")`: how tables and lists are read.
* `vignette("limits-and-encoding")`: resource limits, encodings, and what
  zuhtml does not do.
