---
title: "Real-world pages: long tables, lists and text"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The other guides use small, made-up snippets. This one works on real,
plain, old-style pages: CRAN's own package lists and mirror list, the R
FAQ, Project Gutenberg's top-100 lists, and the directory listings web
servers write. They are the kind of page
zuhtml is built for: large, table- and list-shaped, and written long
before anyone thought of them as data.

This article is built with the package website only, because it fetches
live pages. The numbers in it are from the day the site was last built.

```{r}
library(zuhtml)
```

zuhtml never downloads anything. Each page is fetched once into a
temporary file, and `html_read()` parses the file:

```{r}
fetch <- function(url) {
  path <- tempfile(fileext = ".html")
  utils::download.file(url, path, quiet = TRUE, mode = "wb")
  path
}
cran <- "https://cran.r-project.org/"
```

## A long table: every CRAN package

`available_packages_by_name.html` is one table with a row per package,
about 25,000 rows and several megabytes of HTML.

```{r}
by_name <- html_read(
  fetch(paste0(cran, "web/packages/available_packages_by_name.html")),
  base_url = paste0(cran, "web/packages/available_packages_by_name.html")
)
html_info(by_name)[c("nodes", "input_bytes", "native_bytes")]

packages <- html_tables(by_name)[[1]]
dim(packages)
head(packages)
```

The table has no header row, so the columns are `V1` and `V2`, and a few
rows are empty: they are the targets of the A-to-Z index at the top of
the page. The markup says so, which makes a more precise extraction
possible. The index rows carry an `id`, the package rows do not, and each
package row has a link:

```{r}
rows <- html_elements(by_name, "table tr:not([id])")
length(rows)

packages <- data.frame(
  package = html_text_clean(html_element(rows, "td:first-child")),
  title = html_text_clean(html_element(rows, "td:nth-child(2)")),
  url = html_url(html_element(rows, "a"))
)
head(packages, 3)
```

`html_element()` returns one result per row, so the three columns line
up even if some row lacked a link. `html_url()` resolves the relative
`../../web/packages/...` links against the page's address.

With a data frame, the rest is ordinary R:

```{r}
first <- toupper(substr(packages$package, 1, 1))
head(sort(table(first), decreasing = TRUE))
head(packages$package[grepl("\\bshiny\\b", packages$title, ignore.case = TRUE)])
```

## A table with a header: packages by publication date

The same list sorted by date has a header row, so `html_table()` names
the columns itself:

```{r}
by_date <- html_read(
  fetch(paste0(cran, "web/packages/available_packages_by_date.html"))
)
recent <- html_tables(by_date)[[1]]
names(recent)
head(recent, 3)

year <- substr(recent$Date, 1, 4)
tail(table(year), 8)
```

Dates stay character, as every cell does: convert them when you know
they are dates.

```{r}
range(as.Date(recent$Date))
```

## Many small tables: CRAN mirrors

The mirror list is a definition list: each `<dt>` names a country, and
the `<dd>` after it holds a table of that country's mirrors.

```{r}
mirrors_page <- html_read(fetch(paste0(cran, "mirrors.html")))
countries <- html_elements(mirrors_page, "dt")
length(countries)
head(html_text_clean(countries))
```

`html_next_sibling()` is aligned, one result per `<dt>`, so each
country's table can be read beside its name:

```{r}
mirrors <- do.call(rbind, lapply(countries, function(dt) {
  tab <- html_element(html_next_sibling(dt), "table")
  rows <- html_table(tab, header = FALSE)
  data.frame(country = html_text_clean(dt), url = rows$V1, host = rows$V2)
}))
nrow(mirrors)
head(mirrors[, c("country", "url")])
sort(table(mirrors$country), decreasing = TRUE)[1:5]
```

## Lists of strings: Project Gutenberg's top 100

Project Gutenberg's top-100 page has six ordered lists, each under a
heading:

```{r}
top <- html_read(fetch("https://www.gutenberg.org/browse/scores/top"))
html_text_clean(html_elements(top, "h2"))
```

`html_list()` reads one list into a character vector, one string per
item. The lists are the `<ol>`s right after a heading, and
`html_previous_sibling()`, being aligned, gives each list's heading:

```{r}
ols <- html_elements(top, "h2 + ol")
lists <- lapply(ols, html_list)
names(lists) <- html_text_clean(html_previous_sibling(ols))
lengths(lists)
head(lists[["Top 100 EBooks yesterday"]])
```

Each item is a string of the form "Title by Author (downloads)". Plain
string handling turns a list into a data frame:

```{r}
books <- lists[["Top 100 EBooks last 30 days"]]
m <- regmatches(books, regexec("^(.*) by (.*) \\(([0-9]+)\\)$", books))
ok <- lengths(m) == 4
top30 <- data.frame(
  title = vapply(m[ok], `[`, "", 2),
  author = vapply(m[ok], `[`, "", 3),
  downloads = as.integer(vapply(m[ok], `[`, "", 4))
)
head(top30)
```

A few items have no author, and the pattern skips them (`r sum(!ok)`
this time).

## Text: the R FAQ

The R FAQ is a long page of prose, lists and code. Its headings are the
table of contents:

```{r}
faq <- html_read(fetch(paste0(cran, "doc/FAQ/R-FAQ.html")))
headings <- html_text_clean(html_elements(faq, "h2, h3"))
length(headings)
head(sub(" \u00b6$", "", headings), 8)
```

(The `¶` is the text of the permalink next to each heading.)

Each numbered section is a `div` holding its heading, its paragraphs and
a navigation bar, which is itself a paragraph, `<p class="nav-panel">`.
Reading every section into a list of paragraph strings takes one aligned
call for the titles and one `lapply()` for the bodies.
`:scope > p:not(.nav-panel)` keeps the section's own paragraphs, not
those of nested elements, and leaves out the navigation bar:

```{r}
sections <- html_elements(faq, "div.section-level-extent")
titles <- sub(" \u00b6$", "", html_text_clean(html_element(sections, "h3")))
paragraphs <- lapply(sections, function(s) {
  html_text_clean(html_elements(s, ":scope > p:not(.nav-panel)"))
})
names(paragraphs) <- titles
length(paragraphs)
paragraphs[["2.1 What is R?"]]
```

`html_text_clean()` collapses whitespace and drops markup, which suits
prose. For code, `html_text()` keeps the text exactly as written:

```{r}
code <- html_elements(faq, "pre.example-preformatted")
length(code)
cat(html_text(code[[1]]))
```

Lists inside the prose come out one string per item:

```{r}
bullets <- html_elements(faq, "div.section-level-extent ul")
length(bullets)
html_list(bullets[[1]])
```

To keep a section's structure (headings, lists, code and links) rather
than just its text, `html_markdown()` writes it as Markdown. Links are
resolved against the page's address, which `html_read()` does not know,
so the page is parsed again with a `base_url`. The navigation bar and
the rule under it are left out by selecting the section's other
children:

```{r}
faq_url <- paste0(cran, "doc/FAQ/R-FAQ.html")
faq <- html_read(fetch(faq_url), base_url = faq_url)
sections <- html_elements(faq, "div.section-level-extent")
s <- sections[startsWith(html_text_clean(html_element(sections, "h3")),
                         "2.7 ")]
parts <- html_elements(s, ":scope > :not(.nav-panel):not(hr)")
cat(head(html_markdown(parts), 7), sep = "\n\n")
```

The Markdown renders back to the same structure. For example, with the
commonmark package, which zuhtml does not depend on:

```{r, eval = requireNamespace("commonmark", quietly = TRUE)}
back <- html_parse(commonmark::markdown_html(
  paste(html_markdown(parts), collapse = "\n\n")
))
html_text_clean(html_elements(back, "h3"))
length(html_elements(back, "li"))
length(html_elements(back, "pre"))
```

## Directory listings

Web servers that list a directory's files write the oldest-style pages
of all: a title "Index of /path", a link to the parent directory at the
top, then one line per file. There are two common layouts, and both are
easy to read with the right tool.

### Listings as tables

Apache's fancy index, used by CRAN and by GNU's mirror, is a table: a
header row, a row holding only a rule (`<hr>`), then one row per entry,
the first being "Parent Directory".

```{r}
r4_url <- paste0(cran, "src/base/R-4/")
r4 <- html_read(fetch(r4_url), base_url = r4_url)
html_text_clean(html_element(r4, "title"))

listing <- html_tables(r4)[[1]]
head(listing, 3)
```

Both leading rows are all `<th>` cells, so `html_table()` takes them as
the header; the rule row is empty and adds nothing to the names. The
first column holds only an icon. Its `alt` text says what each entry is,
which the table's cleaned text cannot see, so read it separately,
starting from the link in each entry's second cell:

```{r}
entries <- html_elements(r4, "tr td:nth-child(2) a")
kind <- html_attr(html_element(html_parent(html_parent(entries)), "img"), "alt")
table(kind)

files <- data.frame(
  name = html_text(entries),
  url = html_url(entries),
  kind = kind
)
files <- files[files$kind != "[PARENTDIR]", ]
tail(files, 3)
```

The parent-directory link is an absolute path, `/src/base/`, and
`html_url()` resolves it against the listing's address, so walking up a
level is one more `html_read()`:

```{r}
parent_url <- html_url(html_element(r4, "a:not([href^='?'])"))
parent_url
parent <- html_read(fetch(parent_url), base_url = parent_url)
# Subdirectories end in "/"; the parent link is absolute, the rest are not.
subdirs <- html_elements(
  parent, "tr td:nth-child(2) a[href$='/']:not([href^='/'])"
)
html_text(subdirs)
```

The sizes (`32M`) are text. A small helper turns the suffixes into
bytes:

```{r}
size_bytes <- function(x) {
  x <- trimws(x)
  mult <- c(K = 2^10, M = 2^20, G = 2^30)[substring(x, nchar(x))]
  num <- suppressWarnings(as.numeric(sub("[KMG]$", "", x)))
  ifelse(is.na(mult), num, num * mult)
}
listing <- listing[grepl("tar\\.gz$", listing$Name), ]
listing$bytes <- size_bytes(listing$Size)
head(listing[, c("Name", "Last modified", "bytes")], 3)
```

### Listings as preformatted text

nginx, which serves kernel.org, writes the listing inside one `<pre>`:
a link, then plain text with the date and size, then a newline, for
each file.

```
<pre><a href="../">../</a>
<a href="ChangeLog-6.0">ChangeLog-6.0</a>        03-Oct-2022 05:09     14M
```

There is no cell to select for the date. It is the text node after each
link, and `html_next_sibling(..., elements_only = FALSE)` returns exactly
that, one per link, aligned:

```{r}
kernel_url <- "https://cdn.kernel.org/pub/linux/kernel/v6.x/"
kernel <- html_read(fetch(kernel_url), base_url = kernel_url)
links <- html_elements(kernel, "pre > a:not([href='../'])")
length(links)

after <- html_text(html_next_sibling(links, elements_only = FALSE))
fields <- strsplit(trimws(after), "[[:space:]]+")
head(fields, 3)

is_file <- lengths(fields) == 3
releases <- data.frame(
  name = html_text(links)[is_file],
  date = vapply(fields[is_file], `[`, "", 1),
  size = size_bytes(vapply(fields[is_file], `[`, "", 3)),
  url = html_url(links[is_file])
)
tarballs <- releases[grepl("^linux-6\\.[0-9.]+\\.tar\\.xz$", releases$name), ]
nrow(tarballs)
tail(tarballs[, c("name", "date", "size")], 3)
```

The dates are written like `03-Oct-2022`; month names are English
whatever the locale, so parse them in the C locale:

```{r}
old <- Sys.getlocale("LC_TIME")
invisible(Sys.setlocale("LC_TIME", "C"))
tarballs$date <- as.Date(tarballs$date, format = "%d-%b-%Y")
invisible(Sys.setlocale("LC_TIME", old))
range(tarballs$date)
```

Apache writes the same preformatted layout when fancy indexing is on
but tables are off, as on archive.apache.org. The header links there
sort the listing (`?C=N;O=D`), and an attribute selector leaves them
out:

```{r}
httpd_url <- "https://archive.apache.org/dist/httpd/"
httpd <- html_read(fetch(httpd_url), base_url = httpd_url)
all_links <- html_elements(httpd, "pre a")
head(html_text(all_links), 5)
entries <- html_elements(httpd, "pre a:not([href^='?'])")
entries <- entries[html_text(entries) != "Parent Directory"]
head(html_text(entries[grepl("/$", html_attr(entries, "href"))]))
```

## A very large page, and the limits

CRAN's check summary, with one row per package and one column per check
flavor, is tens of megabytes of HTML: larger than the default
`max_input` of 16 MiB.

```{r}
summary_page <- fetch(paste0(cran, "web/checks/check_summary_by_package.html"))
file.size(summary_page) / 2^20
err <- tryCatch(html_read(summary_page), zuhtml_limit_error = function(e) e)
err$limit
```

Limits are per call, so a trusted page can simply be given more room:

```{r}
big <- html_limits(max_input = 128 * 2^20, max_memory = 4 * 2^30)
checks <- html_read(summary_page, limits = big)
html_info(checks)[c("nodes", "native_bytes", "parse_peak_bytes")]
status <- html_tables(checks, limits = big)[[1]]
dim(status)
names(status)[1:4]
```

The flavor names contain line breaks because the header cells do: they
are written `r-devel<br>Linux<br>...`, and cleaned text turns `<br>`
into `"\n"`. Replace them if single-line names suit better:

```{r}
names(status) <- gsub("\n", " ", names(status))
names(status)[1:4]
table(status[[3]])[1:5]
```
